---
title: "Case Study: Physical Activity (PA)"
author: "Beniamino Hadj-Amar and Jack Jewson"
date: "04/06/2020"
output: html_document
---


---

This R markdown file is part of the supplementary material to the paper 

      'Bayesian Approximations to Hidden Semi-Markov Models' (2020)
                by B.Hadj-Amar, J.Jewson and M.Fiecas. 

Here, we analyze the physical activity (PA) time series that was invest-
igated using a frequentist HMM by Huang et al. (2018). We seek to condu-
cta similar study but within a Bayesian framework and considering the 
extra flexibility afforded by our proposed methodology to investigate de-
partures from the HMM. In this R markdown file we run the full estimati-
on algorithm for HMM and our HSMM approximation with negative-binomial 
durations. 

The data represent 4 days of 5-min averaged PA recording of a healthy 
subject, for a total of 1150 data points.

---


## Set up

```{r set_up, include=TRUE, echo=FALSE, eval=TRUE, cache=TRUE, results='hide'}

library("miceadds")
library("rstan")
library("bayesplot")
library("bridgesampling")
library("matrixStats")
library("lubridate")

# (your current directory must contain both /stan and /include)
setwd("/Users/beniamino/Desktop/HSMM_Project/Final/")
source.all("include/")

rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```

## Data & Pre-processing

```{r data, include=TRUE, echo=FALSE, eval=TRUE, cache=TRUE, results='hide'}
RA.data <- read.csv(paste0(getwd(), "/data/S16.csv"))

# --- NAs (mean adjecent point)
NAs <- which(is.na(RA.data$activity)) # only 3 NA's
for (i in 1:length(NAs)) {
  ind_lower <- NAs[i] - 1
  ind_upper <- NAs[i] + 1
  RA.data$activity[NAs[i]] <- mean(RA.data$activity[c(ind_lower, ind_upper)])
}

# -- transformed data (sqrt)
RA.data$activity.sq <- sqrt(RA.data$activity) 

attach(RA.data)
obs <- activity.sq
N <- nrow(RA.data)
```

## Plot: PA and sqrt(PA)

```{r plot_data, include=TRUE, echo=FALSE, eval=TRUE, cache=TRUE, results='hide'}
# physical activity 
ticks <- seq(from = 3, to = N, by = 96)
labels <- paste(hour(as.character(time))[ticks], "00", sep = ":")
labels[labels == "4:00"] = "04:00"
par(mfrow = c(1, 1))
plot(activity, type = "p", pch = 20, cex = 0.4, xaxt = "n", 
     xlab = "", col = "gray48", xlim = c(-5, N+5), xaxs='i', 
     ylab = "Physical Activity")
axis(side = 1, ticks, labels, cex.axis = 0.7, las = 2, tck=-0.04)
i <- c(0, 2*144, 4*144, 6*144, 8*144)
rect(3 + i, -2, 147 + i, -1, col="black") 

# sqrt physical activity 
ticks <- seq(from = 3, to = N, by = 96)
labels <- paste(hour(as.character(time))[ticks], "00", sep = ":")
labels[labels == "4:00"] = "04:00"
par(mfrow = c(1, 1))
plot(sqrt(activity), type = "p", pch = 20, cex = 0.4, xaxt = "n", 
     xlab = "", col = "gray48", xlim = c(-5, N+5), xaxs='i', 
     ylab = "Sqrt (Physical Activity)", ylim = c(-0.06, 7.2))
axis(side = 1, ticks, labels, cex.axis = 0.7, las = 2, tck=-0.04)
i <- c(0, 2*144, 4*144, 6*144, 8*144)
rect(3 + i, -0.3, 147 + i, -0.2, col="black") 

```

## Comparable Priors - Set Up

```{r prior_setup, include=TRUE, echo=FALSE, eval=TRUE, cache=TRUE, results='hide'}
K <- 3 # n of states
alpha_0 <- matrix(NA, K, K) # hyperparms dirichlet
a_0 <- numeric(K) # hyperparms gamma
b_0 <- numeric(K) #       ""

```


## Comparable Priors - State 1 : Inactive (IA) 

```{r prior_state1, include=TRUE, echo=FALSE, eval=TRUE, cache=TRUE, results='hide'}
# mean 7.5x12 = 90; var (0.5x12)^2 = 36. 
# prior information: we assume that the night rest period of a healhty
# individual is between 7 and 8 hours, where we account for some 
# variability amongst people.

state <- 1
mean.target <- 7.5*12
var.target <- (0.5*12)^2

dir.parms <- optim(par = rep(5, 2),
                   fn = function(par) {
                     geom.dwell.error(exp(par[1]), exp(par[2]), 
                                      mean.target, var.target)
                   })
# we assume equal probability of going to state 2 (MA) or 3 (HA)
est <- exp(dir.parms$par)
alpha_0[state, ] <- c(est[1], est[2]/2, est[2]/2)

gamma.parms <- optim(par = rep(5, 2), 
                     fn = function(par) {
                       nb.dwell.error(exp(par[1]), exp(par[2]), 
                                      mean.target, var.target)
                     })
est <- exp(gamma.parms$par)
a_0[state] <- est[1]; 
b_0[state] <- est[2];


geom.dwell.info(alpha_0[state, state], sum(alpha_0[state, -state]))
nb.dwell.info(a_0[state], b_0[state])
```


## Comparable Priors - State 2 : Moderately Active (MA) 

```{r prior_state2, include=TRUE, echo=FALSE, eval=TRUE, cache=TRUE, results='hide'}
# mean: 1x12 = 12; var: (1.5x12)^2 = 324. 
# prior information:  the other 16-17 hours are split between the other two states, i.e.
# moderately active and highly active states. In general, we have much less information 
# on how long peopel spend in  these states. We fix mean 2 hours and a large variance.
state <- 2
mean.target <- 2*12
var.target <- (1.5*12)^2

dir.parms <- optim(par = c(5, 2),
                   fn = function(par) {
                     geom.dwell.error(exp(par[1]), exp(par[2]), 
                                      mean.target, var.target)
                   })
# MA and HA alternates between each other more frequently than with IA, 
# Hence we may assume transitions from MA->HA is 4 times more likely 
# than going from MA->IA
est <- exp(dir.parms$par)
alpha_0[state, ] <- c(est[2]/5, est[1], est[2]*(4/5))

gamma.parms <- optim(par = c(5, 5), 
                     fn = function(par) {
                       nb.dwell.error(exp(par[1]), exp(par[2]), 
                                      mean.target, var.target)
                     })
est <- exp(gamma.parms$par)
a_0[state] <- est[1]; 
b_0[state] <- est[2];


geom.dwell.info(alpha_0[state, state], sum(alpha_0[state, -state]))
nb.dwell.info(a_0[state], b_0[state])
```


## Comparable Priors - ---- State 3:  Highly Active (HA) 

```{r prior_state3, include=TRUE, echo=FALSE, eval=TRUE, cache=TRUE, results='hide'}
# mean: 1x12 = 12; var: (1.5x12)^2 = 324. 
# prior information:  the other 16-17 hours are split between the other two states, i.e.
# MA and HA states. In general, we have much less information 
# on how long peopel spend in these states. We fix mean 2 hours and a large variance.

state <- 3
mean.target <- 2*12
var.target <- (1.5*12)^2

dir.parms <- optim(par = rep(5, 2),
                   fn = function(par) {
                     geom.dwell.error(exp(par[1]), exp(par[2]), 
                                      mean.target, var.target)
                   })

# MA and HA alternates between each other more frequently than with IA, 
# Hence we may assume transitions from HA->MA is 4 times more likely 
# than going from HA->IA
est <- exp(dir.parms$par)
alpha_0[state, ] <- c(est[2]/5, est[2]*(4/5),est[1])

gamma.parms <- optim(par = rep(5, 2), 
                     fn = function(par) {
                       nb.dwell.error(exp(par[1]), exp(par[2]), 
                                      mean.target, var.target)
                     })
est <- exp(gamma.parms$par)
a_0[state] <- est[1]; 
b_0[state] <- est[2];


geom.dwell.info(alpha_0[state, state], sum(alpha_0[state, -state]))
nb.dwell.info(a_0[state], b_0[state])
```

## Bayesian HSMM Approx - Negative Binomial Dwell

```{r prior_specification, include=TRUE, echo=FALSE, eval=TRUE, cache=TRUE, results='hide'}
K <- 3 # n of states
m <- c(150, 10, 10) # dwell threshold

# hyperparms
sigma_0 <- 2
mu_0 <- rep(mean(obs), K)
a_0; b_0
v_0 <- matrix(c(alpha_0[1, 2:3], alpha_0[2, c(1, 3)], alpha_0[3, 2:3]),
                  nrow = K, ncol = K -1, byrow = TRUE)

# Testing for sparsity
if((K / sum(m)) < 0.1) {
  path.stan <- "stan/bayesHSMMapprox_GaussEmis_NegBinomDur_sparse.stan"
} else{
  path.stan <- "stan/bayesHSMMapprox_GaussEmis_NegBinomDur.stan"
}

HSMM.data <- list(N = length(obs), K = K, y = obs, m = m, mu_0 = mu_0,
                  sigma_0 = sigma_0, a_0_lambda = a_0, b_0_lambda = b_0,
                  a_0_phi = 2, b_0_phi = 2, alpha_0 = v_0)
HSMM.stan <- stan(file = path.stan, data = HSMM.data,
                  warmup = 1000, iter=(1+5)*1000, chains=1, cores=1)

```

## Diagnostic: Bayesian HSMM Approx - Negative Binomial Dwell

```{r bayesHSMMapprox, include=TRUE, echo=FALSE, eval=TRUE, cache=TRUE, results='hide'}
# --- extracting samples 
sims <- extract(HSMM.stan)
lambda_sims <- sims$lambda
phi_sims <- 1/sims$phim1
mu_sims <- sims$mu
sigma_sims <- sqrt(sims$sigma2)
gamma_sims <- get.gamma_sims(sims$gamma)

# --- bayes estimates
lambda.hat <- colMeans(lambda_sims)
phi.hat <- colMeans(phi_sims)
mu.hat <- colMeans(mu_sims)
sigma.hat <- colMeans(sigma_sims)
gamma.hat <- apply(gamma_sims, c(1, 2), mean)

# --- posterior predictive check
negBinom.HSMM.predictive.plot(sims, obs, m, ndraw = 50)

# --- psuedo-residuals
negBinom.HSMM.pseudo_residuals(obs, m, lambda.hat, phi.hat,
                               mu.hat, sigma.hat, gamma.hat)

```



## Bayesian HMM - Geometric Dwell

```{r bayes_HMM, include=TRUE, echo=FALSE, eval=TRUE, cache=TRUE, results='hide'}
K <- 3 # n of states

# hyperparams
mu_0 <- rep(mean(obs), K)
sigma_0 <- 2
alpha_0

HMM.data <- list(N = length(obs), K = K, y = obs, mu_0 = mu_0,
                 sigma_0 = sigma_0, alpha_0 = alpha_0)
HMM.stan <- stan(file = "stan/bayesHMM_GaussEmis.stan", data = HMM.data,
                 warmup = 1000, iter=(1+5)*1000, chains=1, cores=1)
print(HMM.stan, probs = c(0.05, 0.95))
```



## Diagnostic: Bayesian HMM (Geometric Dwell)

```{r bayes_HMM, include=TRUE, echo=FALSE, eval=TRUE, cache=TRUE, results='hide'}
# --- extracting samples 
sims <- extract(HMM.stan)
mu_sims <- sims$mu
sigma_sims <- sqrt(sims$sigma)
gamma_sims <- sims$gamma

# --- bayes estimates
mu.hat <- colMeans(mu_sims)
sigma.hat <- colMeans(sigma_sims)
gamma.hat <- apply(gamma_sims, c(2, 3), mean)

# psuedo residuals + most likely state sequence (viterbi)
HMM.pseudo_residuals(obs, mu.hat, sigma.hat, gamma.hat, plt = TRUE)
HMM.viterbi(obs, mu.hat, sigma.hat,gamma.hat, draw = FALSE, plt = TRUE) 
```


## Model Selection - Bayes Factor

```{r model_selection, include=TRUE,echo=TRUE, eval=TRUE,cache=TRUE,results='hide'}
# - bridge sampler and marginal likelihood
HSMM.bridge <- bridge_sampler(HSMM.stan)
HMM.bridge <- bridge_sampler(HMM.stan)
HSMM.marg_llk <- HSMM.bridge$logml
HMM.marg_llk <- HMM.bridge$logml

# - bayes factor
exp(HSMM.marg_llk - HMM.marg_llk)
```
