---
title: "BayesApproxHSMM - Tutorial"
author: "Beniamino Hadj-Amar and Jack Jewson"
date: "03/06/2020"
output: html_document
---


---
This R markdown file is part of the supplementary material to the paper 

      'Bayesian Approximations to Hidden Semi-Markov Models' (2020)
                by B.Hadj-Amar, J.Jewson and M.Fiecas. 

Here, we generate data from a three state HSMM with Gaussian emission 
and Poisson dwell durations.Estimation procedures are carried out us-
ing both frequentist (EM) and Bayesian approaches, where we validate 
each model using several diagnostic tools. We also illustrate the use 
of Bayes factors for model selection between the proposed approach and
the HMM.
---



## Set up

```{r set_up, include=TRUE, echo=FALSE, eval=TRUE, cache=TRUE, results='hide'}

library("miceadds")
library("rstan")
library("bayesplot")
library("bridgesampling")
library("matrixStats")
library("lubridate")

# (your current directory must contain both /stan and /include)
setwd("/Users/beniamino/Desktop/HSMM_Project/Final/")
source.all("include/")

rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```


# --------------- Data Generation 

```{r data_generation, include=TRUE,echo=TRUE, eval=TRUE,cache=TRUE,results='hide'}

# -- Parameterization

N <- 200  # length time series
parms <- list()
parms$K <- 3  # n. states
parms$lambda <- c(20, 30, 20) # poisson duration rate
parms$mu <- c(5, 14, 30)  # emission parameter - mean
parms$sigma <- c(1, 1, 1) # emission parameters - sd
parms$gamma <- matrix(c(0, 0.3, 0.7,
                        0.2, 0, 0.8,
                        0.1, 0.9, 0), parms$K, parms$K, byrow = T) #  t.p.m
parms$delta <- c(1.0, 0.0, 0.0) # initial dstr

# - Simulate data

set.seed(5)
simul <- gauss.HSMM.generate_sample(N, parms)
obs <- simul$obs
state <- simul$state
plot(obs, col = state, cex = 0.7, pch = 20, type = "o")

```




# ------------ HSSM approx - Expectation/Conditional Maximization (ECM)

```{r ECM, include=TRUE, echo=TRUE, eval=TRUE,cache=TRUE,results='hide'}

K <- 3 # n states
m <- rep(5, K) # dwell threshold
lambda.0 <- rep(5, K) # lambda initial value 
parms.0 <- HSMM.init(obs, K, lambda.0) # emission initial values
# (different initial values may be required for convergence)
HSMM.ECM.fit <- HSMM.ECM(K = K, m = m, obs = obs, parms_init = parms.0, niter = 1e2)
cat("mllk:", HSMM.ECM.fit$mllk, " AIC:", HSMM.ECM.fit$AIC, " BIC:", HSMM.ECM.fit$BIC, "\n")
```

```{r ECM_out, include=TRUE, echo=TRUE, eval=TRUE, cache=TRUE, results='hide'}
# psuedo residuals + most likely state sequence (viterbi)
HSMM.pseudo_residuals(obs, m, HSMM.ECM.fit$lambda, 
                      HSMM.ECM.fit$mu, HSMM.ECM.fit$sigma, 
                      HSMM.ECM.fit$gamma, plt = TRUE)
HSMM.viterbi(obs, m, HSMM.ECM.fit$lambda, HSMM.ECM.fit$mu, HSMM.ECM.fit$sigma,
             HSMM.ECM.fit$gamma, plt = TRUE)
```


# --------------- HSMM approx -  Bayesian Model 

```{r Bayesian_hsmm_hmm_fit, include=TRUE,echo=TRUE, eval=TRUE,cache=TRUE,results='hide'}
K <- 3 # n states 
m <- rep(5, K) # dwell threshold
lambda.0 <- rep(10, K) # lambda initial value MCMC
data.stan <- list(N = length(obs), K = K, y = obs,
                  m = m,  mu_0 = rep(mean(obs), K), 
                  sigma_0 = 2, a_0 = rep(0.01, K), b_0 = rep(0.01, K),
                  alpha_0 = matrix(1, nrow = K, ncol = K-1))

if ((K / sum(m) < 0.1)) {
  stan_path <- "stan/bayesHSMMapprox_GaussEmis_PoissDur.stan"
} else {
  stan_path <- "stan/bayesHSMMapprox_GaussEmis_PoissDur_sparse.stan"
}
HSMM.stan <- stan(file = stan_path, data = data.stan, 
                 init = function(){HSMM.init.stan(K, obs, lambda.0)}, 
                 warmup = 1000, chains = 1, iter = (1+5)*1000, cores = 1, 
                 control = list(adapt_delta=0.99,stepsize=0.01,max_treedepth = 20))
print(HSMM.stan, probs = c(0.05, 0.95))
```

```{r bayes_out, include=TRUE, echo=TRUE, eval=TRUE,cache=TRUE,results='hide'}

# --- extracting samples 
sims <- extract(HSMM.stan)
lambda_sims <- sims$lambda
mu_sims <- sims$mu
sigma_sims <- sqrt(sims$sigma2)
gamma_sims <- get.gamma_sims(sims$gamma)

# --- bayes estimates
lambda.hat <- colMeans(lambda_sims)
mu.hat <- colMeans(mu_sims)
sigma.hat <- colMeans(sigma_sims)
gamma.hat <- apply(gamma_sims, c(1, 2), mean)
```


```{r bayes_diag, include=TRUE, echo=TRUE, eval=TRUE,cache=TRUE,results='hide'}
# --- trace plots + hist transitions
mcmc_trace(as.array(HSMM.stan), regex_pars = "gamma")
mcmc_trace(as.array(HSMM.stan), regex_pars = "mu")
mcmc_trace(as.array(HSMM.stan), regex_pars = "sigma2")
mcmc_trace(as.array(HSMM.stan), regex_pars = "lambda")
HSMM.transitions.hist(gamma_sims)
```



```{r bayes_diag, include=TRUE, echo=TRUE, eval=TRUE,cache=TRUE,results='hide'}
# psuedo residuals + most likely state sequence (viterbi)
HSMM.pseudo_residuals(obs, m, lambda.hat, mu.hat, sigma.hat, gamma.hat, plt = TRUE)
HSMM.viterbi(obs, m, lambda.hat, mu.hat, sigma.hat,gamma.hat, draw = FALSE, plt = TRUE) 
```


```{r bayes_diag, include=TRUE, echo=TRUE, eval=TRUE,cache=TRUE,results='hide'}
# -- posterior predictive plot check
HSMM.predictive.plot(sims, obs, m, ndraw = 50)
```


```{r bayes_diag, include=TRUE, echo=TRUE, eval=TRUE,cache=TRUE,results='hide'}
# --- predictive performances
HSMM.performance <- HSMM.stan.performance(sims, obs, m)
HSMM.performance
```


# ------------ HMM standard  - Bayes  

```{r bayes_mu, include=TRUE, echo=TRUE, eval=TRUE,cache=TRUE,results='hide'}

K <- 3
HMM.data <- list(N = length(obs), K = K, y = obs, 
                 mu_0 = rep(mean(obs), K), sigma_0 = 2, 
                 alpha_0 = matrix(c(rep(1, K*K)), nrow = K, ncol = K, byrow = TRUE))
HMM.stan <- stan(file = "stan/bayesHMM_GaussEmis.stan", data = HMM.data,
                 warmup = 1000, iter=(1+5)*1000, chains=1, cores=1)
print(HMM.stan, probs = c(0.05, 0.95))
```


# ------------ Model Selection, Bayes Factors ------

```{r bridge, include=TRUE,echo=TRUE, eval=TRUE,cache=TRUE,results='hide'}
# bridge sampler
HSMM.bridge <- bridge_sampler(HSMM.stan)
HMM.bridge <- bridge_sampler(HMM.stan)
```

```{r Bayes_Factor, include=TRUE,echo=TRUE, eval=TRUE,cache=TRUE,results='hide'}
# bayes factor
HSMM.marg_llk <- HSMM.bridge$logml
HMM.marg_llk <- HMM.bridge$logml
exp(HSMM.marg_llk - HMM.marg_llk)
```


